Sediment transport capacity response to variations in water discharge in pressurized subglacial channels

### 2025-02-17 revised_v1/response_R2.tex

\item  \textbf{And the model should come with some caveats, e. g., you could say that it is a simplistic and possibly unreliable model. To be fair, this is done in section 5.3, but that is too late.}

  \textit{The ``Methods'' section is largely meant to describe the model, as opposed to discussing its behavior or simplicity. In response to this comment, we will move the ``Experiment limitations'' to a section directly after the ``Results'' but before the ``Discussion.''}

\item \textbf{  One of the comments made is that because the sub-aerial model has an algebraic relation between width and discharge (equation 9), the relationship between velocity and discharge is also algebraic; but this is simply a choice of the model. In reality, the width of a channel will also evolve over (long) time through processes of bank erosion, and the use of equation 9 properly only involves a long term time average of discharge, so it is misleading to use it in a time-specific way. Actually, this is admitted at line 370, along with other limitations in the models.}

  \textit{We appreciate these comments. With regard to the first point about channel width evolution in the subaerial model, this matter is confronted in Sections 4.2 and 4.3.
    Here we show that across a range of exponents and relationships between channel cross-section and water discharge, subglacial systems are still more sensitive to water discharge variations than subaerial ones.
    This increased variability in subglacial sediment transport capacity occurs across a range of $\alpha$ values that describe the relationship between channel width and water discharge as shown in Figure 6 (will be 7).
    As a result, we believe that these results hold across a range of water discharge- channel width relationships, including ones where $\alpha$ evolves in time.
    The exception, shown in Section 4.3, is that when water discharge evolves more slowly than channel size. Here, the behavior between subaerial and subglacial is very similar, because channel areas in both channel types adjust to discharge variations.}

@@

\item   \textbf{Two kinds of numerical experiments are described in section 3.3, and the results of these presented in section 4. Presumably the input to the model runs is the (time- varying) water discharge, but that seems not to be stated in 3.3, which I would expect, although it is stated in the caption to figure 2.
    The second half of the paper characterises the results of these experiments and there is an extended discussion. There is no doubt the authors have put a lot of effort into this, but my interest waned at this point.}

  \textit{To make it clear that water discharge is the input, we will modify the first sentence of Section 3.3 to explicitly state:}
    ``Proglacial discharge records from the Fieschergletscher (scenario \alpine{}) and the Leverett Glacier (scenario \icesheet{}) are used as inputs for the models above.''

\item \textbf{  In summary, this is one of those difficult papers to judge, because the effort in- volved is quite substantial and honestly applied, but the whole philosophy of the approach is, in my view, misguided. This is a coxless boat crew who are rowing up a backwater without a rudder, and have lost the main direction of the stream.}

  \textit{With respect to the last sentence, it is true that the lead author feels this way occasionally. Regardless, we think several valuable and unique findings are within this manuscript.}

@@

\item \textbf{114:} I had no idea what the central angle referred to, but evidently it is the angle subtended at the centre of the circular arc of the channel upper boundary; and then $\beta = 2\alpha$, where $\alpha$ is the contact angle at the channel edge, which seems a more natural quantity to use. Also I checked the algebra of equation (5), and got this formula, but with the factor of two in the numerator, not the denominator, assuming hydraulic diameter is twice hydraulic radius, the latter of which is area/perimeter, so please check this. Incidentally, I did also then check equation (7) and I agree with that, so I do think there is an error in (5).

  \textit{We thank the reviewer for the careful assessment of our work. We have double-checked and believe that there is no problem in the equation. Our proof is below:}

   \includegraphics[width=0.3\linewidth]{Hooke.png}
  Guiding equations:

  Hydraulic diameter:
  \begin{equation}
    D_h = \frac{4S_g}{P_w}~~~(I)
  \end{equation}

  Wetter perimeter of the conduit \citep[][pg. 69]{hooke1990}:

  \begin{equation}
    P_w = R \beta +2R \sin\frac{\beta}{2}= R(\beta +2\sin \frac{\beta}{2}) ~~~(II)
  \end{equation}

  Cross sectional area of the conduit \citep[][pg. 69]{hooke1990}:
  \begin{equation}
    S_g=\frac{R^2}{2}(\beta - \sin \beta)~~~(III)
  \end{equation}

@@

  $D_h$ is hydraulic diameter, $P_w$ is wetter perimeter, $R$ is radius, $\beta$ is Hooke angle \citep{hooke1990}, and $S_g$ is the subglacial channel cross section.

  Proof:

  \begin{linenomath*}
    \begin{equation}
      \frac{D_h}{4S_g} = \frac{1}{R\beta + 2R\sin\frac{\beta}{2}} \quad I \mathrm{into~} II 
    \end{equation}
  \end{linenomath*}

@@

  \begin{equation}
    R = \frac{4S_g}{D_h (\beta +2 \sin \frac{\beta}{2})}, \quad \mathrm{into~} III 
  \end{equation}

  \begin{equation}
    S_g = \frac{16 S_g^2}{D_h^2 (\beta +2\sin\frac{\beta}{2})^2} ~~~\frac{1}{2}(\beta-\sin\beta)
  \end{equation}

  \begin{equation}
    1 = \frac{8 S_g}{D_h^2 (\beta +2\sin\frac{\beta}{2})^2} ~~~ (\beta-\sin\beta)
  \end{equation}

  \begin{equation}
    \frac{D_h^2}{8} ~~~ \frac{(\beta+2\sin\frac{\beta}{2})^2}{\beta - \sin \beta} = S_g
  \end{equation}

@@

  \begin{equation}
    \frac{D_h^2 (2(\frac{\beta}{2}+\sin \frac{\beta}{2}))^2}{8(\beta - \sin \beta)} = S_g
  \end{equation}

   \begin{equation}
    \frac{D_h^2 (\frac{\beta}{2}+\sin \frac{\beta}{2})^2}{2(\beta - \sin \beta)} = S_g
  \end{equation}

### 2025-02-17 Delaneyetal2024.tex
preparing diff file

\subsection{Timing and covariance of subaerial and subglacial  sediment transport capacity variations}

The first numerical experiment aims to quantify the timing and covariance of the subglacial model outputs with respect to water discharge in response to different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases as channel size evolves (Figure~\ref{fig:model_outs} a, e).
Because subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
An inconsistent relationship between subglacial model outputs and water discharge persists across the range of parameters examined in the ensemble runs (see Section~\ref{sect:ensemble}; Figure S15).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.75\linewidth]{Fig2.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with 
  }
  \label{fig:model_outs}
\end{figure}

@@

\FloatBarrier
\section{Discussion}

@@

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different channel types and  $\alpha$ values.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow channel type.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$ for the MPM (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used previously scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section.
Here the cross-section responds to a characteristic discharge, but variations in water discharge deviate substantially from the flow conditions responsible for the channel size \citep[i.e. diurnal water discharge variations in \alpine{} case Figure \ref{fig:Qw_vari}; e.g.][]{gimbert2016}.
Table~\ref{tab:Qs} shows that sediment transport in pipe-flow scales much more severely with discharge; the exponent on $Q$ being between $3$ and $5$, compared to the other two channel types when that exponent is at most $\frac{5}{3}$.
Thus fluctuations of discharge on short timescales (on the order of a day; Figure~\ref{fig:multi_run}) have the potential to cause conditions with very high sediment transport capacities.
Alternatively, at low water discharges these channels could become depressurized and transition to subearial flow \citep{perolo2018}.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep[][]{alley1997}.
Note, however, that pipe flow assumptions would cause sediment discharge capacity to covary with water discharge, as is not the case in normal R-channels (Figures~\ref{fig:Qw_vari}~and~\ref{fig:chan_types}\,a\,and\,b).
Covariance also occurs for steady-state R-channels (Figure \ref{fig:chan_types} d).

\subsection{Increased variability in  sediment transport capacity in subglacial systems}
\label{sect:dis_qsc}

@@

>>>>

\subsection{Timing and covariance of subaerial and subglacial  sediment transport capacity variations}
\label{sect:timing}
The first numerical experiment aims to quantify the timing and covariance of the subglacial model outputs with respect to water discharge in response to different seasonal evolutions and peaks (Figure~\ref{fig:model_outs}).
Variable relationships between model outputs and water discharge emerge for the subaerial and subglacial cases as channel size evolves (Figure~\ref{fig:model_outs} a, e).
Because subaerial channels have no history dependence, each water discharge value in the subaerial channel produces a unique water velocity, shear stress, and width-integrated shear stress (Section~\ref{sect:sub_mode}).
This characteristic results in a perfect rank correlation between the variables and water discharge (Figure~\ref{fig:Qw_vari}).
An inconsistent relationship between subglacial model outputs and water discharge persists across the range of parameters examined in the ensemble runs (see Section~\ref{sect:ensemble}; Figure S15).
\begin{figure}[hbt!]
  \centering
  \includegraphics[width=0.75\linewidth]{Fig2.png}
  \caption{Model outputs from simulations using the hydrographs in panels a and e for the scenarios \alpine{} (a-d) and \icesheet{} (e-h).
    Black (gray) lines in a and e represent the subglacial channel size (water discharge).
    Blue (orange) lines represent outputs from the subglacial (subaerial) channel.
    Data are shown at $15$\unit{min} intervals.
    Arrows show examples where variables peak in the subglacial channel before the subaerial one.
    Insets in f-h show the peak melt event denoted by the shaded area in panels e--h, with 
  }
  \label{fig:model_outs}
\end{figure}

@@

@@

Table~\ref{tab:Qs} shows the scaling of the proxy $w\tau$ as well as $Q_s$ of the MPM, EH, and Bagnold sediment transport formulas with respect to $Q$, $\Psi$ or $S$ for the three different channel types and  $\alpha$ values.
Sediment transport in pipe-flow scales much more severely with discharge (Table~\ref{tab:Qs}); the exponent on $Q$ being between $3$ and $5$, compared to the other two channel types when that exponent is at most $\frac{5}{3}$.
Remarkably, the Bagnold formula has a negative exponent for $f$ in all but the pipe-flow channel type.
The total transport formula EH gives a slightly stronger dependence on all variables due to the larger exponent on $\tau$ of $\frac{5}{2}$ versus  $\frac{3}{2}$ for the MPM (Equation~\ref{eq:Qs_eq}).
Albeit, the sediment transport response in pipe flow for the Bagnold case is close to EH.
Conversely, the sediment transport proxies $w_g\,\tau_g$ and $w_f\,\tau_f$ used previously scale only as $Q^2$ for pipe-flow, whereas the sediment transport capacity scales at least with $Q^3$.
The exponent for the width scaling $\alpha$ only impacts the relationship between sediment transport and water discharge in the EH relation in any meaningful way.
However, for the value of $\alpha$ around $\frac{1}{3}$ (a value appropriate for most streams), the exponent on $Q$ for EH is only slightly greater than the other transport relations.

@@

\FloatBarrier
\section{Discussion}
\subsection{Increased variability in  sediment transport capacity in subglacial systems}
\label{sect:dis_qsc}

@@

R-channels rarely operate in a steady state with variations in water discharge, especially during severe rain or melt is too short for the channel to reach a steady state (Figures~\ref{fig:model_outs} and \ref{fig:Qw_vari}).
In these cases with high water discharge variability, channels can behave more like a pipe of fixed cross-section.
Here the cross-section responds to a characteristic discharge, but variations in water discharge deviate substantially from the flow conditions responsible for the channel size \citep[i.e. diurnal water discharge variations in \alpine{} case Figure \ref{fig:Qw_vari}; e.g.][]{gimbert2016}.
Thus fluctuations of discharge on short timescales (on the order of a day; Figure~\ref{fig:multi_run}) have the potential to cause conditions with very high sediment transport capacities.
Alternatively, at low water discharges these channels could become depressurized and transition to subearial flow \citep{perolo2018}.
These sediment transport capacities variations are of far higher magnitude than those of subaerial channels \citep[][]{alley1997}.
Note, however, that pipe flow assumptions would cause sediment discharge capacity to covary with water discharge, as is not the case in normal R-channels (Figures~\ref{fig:Qw_vari}~and~\ref{fig:chan_types}\,a\,and\,b).
Covariance also occurs for steady-state R-channels (Section \ref{sect:timing}; Figure \ref{fig:chan_types} d).
